BETTING AND BELIEF: PREDICTION MARKETS AND ATTRIBUTION OF CLIMATE CHANGE

### 2016-03-28 README.Rmd
pre-AGU work on time series.

This is code for an agent-based model (ABM) of a climate prediction market in which traders adapt their belief about the climate based on the monetary performance of their neighboring traders in a social network. The ABM is meant to test whether climate prediction markets foster the convergence of people’s beliefs about the climate.

@@

>>>>

# Betting and Belief: Prediction Markets and Attribution of Climate Change

@@

Despite much scientific evidence, a large fraction of the American public doubts that greenhouse gases are causing global warming. We present a simulation model as a computational test-bed for climate prediction markets. Traders adapt their beliefs about future temperatures based on the profits of other traders in their social network. We simulate two alternative climate futures, in which global temperatures are primarily driven either by carbon dioxide or by solar irradiance. These represent, respectively, the scientific consensus and a hypothesis advanced by prominent skeptics. We conduct sensitivity analyses to determine how a variety of factors describing both the market and the physical climate may affect traders' beliefs about the cause of global climate change. Market participation causes most traders to converge quickly toward believing the "true" climate model, suggesting that a climate market could be useful for building public consensus.

This repository contains all the code and data for this simulation model.

Under most parameterizations, within 15-20 years, the median fraction of traders believing in the true model is over 75%:

```{r time, eval=TRUE, echo=FALSE, message=FALSE, fig.width=10.5, fig.height=8.5}
source("utils/multi_plot.R")
source("utils/time_plot.R")
theme_set(theme_classic())
p1 <- history_box() + theme(text=element_text(family='Palatino'))
p2 <- future_box() + theme(text=element_text(family='Palatino'))
multi_plot(p1, p2, cols=2)
```

### 2016-03-28 prediction_market_wsc.tex

The climate change debate has become strongly polarized over the past two decades. Although the scientific consensus on the anthropogenic nature of climate change strongly increased, beliefs about climate change did not evolve much within the public \shortcite{Vandenbergh2013f}. In addition, the divide on anthropogenic climate change between liberals and conservatives has grown steadily as the question is becoming increasingly politicized and potentially disconnected from scientific evidence. The costs of misinformed climate policies are high. If climate change is not human-induced but is believed to be so, public resources will be spent on unnecessary efforts. On the other hand, if climate change is human-induced but not recognized as so, the costs of inaction could be devastating. Given that effective climate policies require long-lasting measures to be taken, the importance of an accurate consensus on the issue is manifest.

Attempts to foster such consensus face many social and psychological challenges, some of which could be adressed by creating climate prediction markets where participants can ``put their money where their mouths are'' \shortcite{Hsu2011,Vandenbergh2013f}. The idea of using prediction markets to efficiently aggregate information about uncertain event outcomes has been widely discussed \shortcite{Horn2014c}. Prediction markets have interesting theoretical properties \shortcite{Set1998,Hanson2012}, and perform well in terms of prediction accuracy and information aggregation in experiments \shortcite{Hanson2006,Healy2010} simulation models \shortcite{Klingert2012c,Jumadinova2011a} and the real-world \shortcite{Wolfers2006,pathak_comparison_2015,Dreber_2015}.

@@

 At each time $t$, traders:
\begin{enumerate}
\item  recalibrate their approximate model based on the new set of temperature data available at $t$,
\label{step:calibrate}
\item  use the posterior probability distribution from (\ref{step:calibrate}) to 
assign beliefs about the probability distribution of future temperatures at time $t^*$: 
$P_{t,m}({T}_{t^*} ~|~ {T}_{0:t}, \text{CO2}_{0:t^*}, \text{TSI}_{0:t^*})$ 
and use it to determine the expected value they attach to each security, and
\item  trade on the CDA market as follows: 
  \begin{itemize}
    \item Every trader $i$ choses at random a security $s_i^B$ she will try to buy.
    \item Every trader $i$ also chooses at random a security $s_i^S$ she will try to sell among the securities she owns a positive amount of (if any).
    \item Traders then decide of their selling price $p_i^S$ and buying price $p_i^B$. 
  To do so, traders first compute their expected values $E(s_i^B)$ and $E(s_i^S)$ for securities $s_i^B$ and $s_i^S$. 
  Then traders set $p_i^S$ at random above $E(s_i^S)$ and $p_i^B$ at random below $E(s_i^B)$ (see Model Parameters below for more details).
    \item Traders go to the market one at the time, in an order drawn randomly for each $t$.
    \item When trader $i$ comes to the market, she places limit orders in the order book. 
   These orders specify that $i$ is willing to buy $s_i^B$ at any price below $p_i^B$, and to sell $s_i^S$ at any price above $p_i^S$.
    \item The market maker then tries to match $i$'s orders with some order which was put in the book before $i$ came to the market.
    \item If there are  outstanding sell offers for $s_i^B$ at a price below $p_i^B$, a trade is concluded. 
   Trader $i$ buys one unit of $s_i^B$ from the seller who sells at the lowest price below $p_i^B$, and the sell and buy offers are removed from the order book.
    \item If there are outstanding buy offers for $s_i^S$ at price above $p_i^S$, a trade is concluded. 
   Trader $i$ sells one unit to the buyer who buys at the highest price above $p_i^S$, and the sell and buy offers are removed from the order book.
    \item Whenever all traders have come to the market, any remaining outstanding offer is removed from the order book, and the trading period is concluded. 
  \end{itemize}
\end{enumerate}

@@

    Formally, let $\texttt{reserv}_{i,t}$ be the reservation price of trader $i$ at time $t$ for some security $s$.
    Then trader $i$ picks her buying or selling price for $s$ uniformly at random in the interval  
    $[(1 - \texttt{risk.tak}_i) \texttt{reserv}_{i,t}, \allowbreak  \texttt{reserv}_{i,t}]$ for buying and 
    $[\texttt{reserv}_{i,t}, \allowbreak (1  + \texttt{risk.tak}_i) \texttt{reserv}_{i,t}]$ for selling. 
  \item  \texttt{ideo} :  determines the degree of ``ideology'' of traders.
    For each trader $i$, the level of $\texttt{ideo}_i$ is drawn uniformly at random from $[0,\texttt{ideo}]$.
    If \texttt{ideo} is high, traders will not revise their approximate models easily, even when faced with evidence that their richest neighbor is doing better than them. 
    Formally, for each trader $i$ and each sequence, a parameter $d_{i}$ is drawn from $[0, \texttt{ideo}_i]$. 
    The value of $d_{i}$ is the probability that $i$ adopts the approximate model of her richest neighbor if that neighbor is doing better than $i$ at the end of the sequence (in monetary terms).
  \end{itemize}
\end{itemize}

@@

Our primary focus is on simulating the market in future scenarios. This is more relevant for policy design and more interesting theoretically due to the increasing divergence of global temperature values under the two future scenarios. The sensitivity analysis is based on a Latin hypercube sampling of parameter sets from the following distributions \shortcite{beachkofski_improved_2002,carnell_lhs_2012}.

@@

We use the model to perform 10,000 simulations for each input paramter set, and then conducted a partial rank correlation coefficient analysis on the relationship between the input matrix, $X$, and the resulting simulated outcome vector of belief convergence scores, $y$ \shortcite{Marino2008,pujol_sensitivity:_2014,saltelli_sensitivity_2009}. Partial correlation computes the linear relationship between the part of the variation of $X_i$ and $y$ that are linearly independent of other $X_j$  ($j \neq i$).  The only difference between the partial correlation and the partial rank correlation that we use here is that the outcome variable $y$ is first ranked-transformed in order to allow us capture potentially non-linear relationships: every value, $y_s$, in the data set is replaced by a number 
$f(y_s) \equiv |\{y_r~|~ y_s > y_r\}|$, 
where $y_r$ are the observed values of $y$ in the data set.

>>>>

The climate change debate has become strongly polarized over the past two decades. Although the scientific consensus on the anthropogenic nature of climate change strongly increased, beliefs about climate change did not evolve much within the public \shortcite{Vandenbergh2013f}. In addition, the divide on anthropogenic climate change between liberals and conservatives has grown steadily as the question is becoming increasingly politicized and potentially disconnected from scientific evidence. The costs of misinformed climate policies are high. If climate change is not human-induced but is believed to be so, public resources will be spent on unnecessary efforts. On the other hand, if climate change is human-induced but not recognized as so, the costs of inaction could be devastating. Given that effective climate policies require long-lasting measures to be taken, the importance of a timely and accurate consensus on the issue is manifest.

Attempts to foster such consensus face many social and psychological challenges, some of which could be addressed by creating climate prediction markets where participants can ``put their money where their mouths are'' \shortcite{Hsu2011,Vandenbergh2013f}. The idea of using prediction markets to efficiently aggregate information about uncertain event outcomes has been widely discussed \shortcite{Horn2014c}. Prediction markets have interesting theoretical properties \shortcite{Set1998,Hanson2012}, and perform well in terms of prediction accuracy and information aggregation in experiments \shortcite{Hanson2006,Healy2010} simulation models \shortcite{Klingert2012c,Jumadinova2011a} and the real-world \shortcite{Wolfers2006,pathak_comparison_2015,Dreber_2015}. However, to the best of our knowledge, the idea that prediction markets can generate consensus \emph{on the factors\/} affecting uncertain events has never been explored quantitatively.

@@

 At each time $t$, traders:
\begin{enumerate}
\item  re-calibrate $m$, their approximate model at time $t$, based on the new set of temperature data available at $t$,
\label{step:calibrate}
\item  use the posterior probability distribution from (\ref{step:calibrate}) to 
assign beliefs about the probability distribution of future temperatures at time $t^*$: 
$P_{t,m}({T}_{t^*} ~|~ {T}_{0:t}, \text{CO2}_{0:t^*}, \text{TSI}_{0:t^*})$ 
and use it to determine the expected value they attach to each security, and
\item  trade on the CDA market as follows: 
  \begin{itemize}
    \item Every trader $i$ chooses at random a security $s_i^B$ she will try to buy.
    \item Every trader $i$ also chooses at random a security $s_i^S$ she will try to sell among the securities she owns a positive amount of (if any).
    \item Traders then decide of their selling price $p_i^S$ and buying price $p_i^B$. 
  To do so, traders first compute their expected values $E(s_i^B)$ and $E(s_i^S)$ for securities $s_i^B$ and $s_i^S$ (where expected values are with respect $i$'s approximate model at time $t$). 
  Then traders set $p_i^S$ at random above $E(s_i^S)$ and $p_i^B$ at random below $E(s_i^B)$ (see Model Parameters below for more details).
    \item Traders go to the market one at the time, in an order drawn randomly for each $t$.
    \item When trader $i$ comes to the market, she places limit orders in the order book. 
   These orders specify that $i$ is willing to buy $s_i^B$ at any price below $p_i^B$, and to sell $s_i^S$ at any price above $p_i^S$.
    \item The market maker then tries to match $i$'s orders with some order which was put in the book before $i$ came to the market.
    \item If there are  outstanding sell offers for $s_i^B$ at a price below $p_i^B$, a trade is concluded. 
   Trader $i$ buys one unit of $s_i^B$ from the seller who sells at the lowest price below $p_i^B$, and the sell and buy offers are removed from the order book.
    \item If there are outstanding buy offers for $s_i^S$ at price above $p_i^S$, a trade is concluded. 
   Trader $i$ sells one unit to the buyer who buys at the highest price above $p_i^S$, and the sell and buy offers are removed from the order book.
    \item Whenever all traders have come to the market, any remaining outstanding offer is removed from the order book, and the trading period is concluded. 
  \end{itemize}
\end{enumerate}

@@

    Formally, in each period, trader $i$ picks her buying or selling price for $s$ uniformly at random in the interval  
    $[(1 - \texttt{risk.tak}_i) \texttt{reserv}_{i}, \allowbreak  \texttt{reserv}_{i,t}]$ for buying and 
    $[\texttt{reserv}_{i}, \allowbreak (1  + \texttt{risk.tak}_i) \texttt{reserv}_{i,t}]$ for selling. 
  \item  \texttt{ideo} :  determines the degree of ``ideology'' of traders.
    For each trader $i$, the level of $\texttt{ideo}_i$ is drawn uniformly at random from $[0,\texttt{ideo}]$.
    If \texttt{ideo} is high, traders will not revise their approximate models easily, even when faced with evidence that their richest neighbor is doing better than them. 
    Formally, for each trader $i$ and each sequence, $\texttt{ideo}_i$ is the probability that $i$ adopts the approximate model of her richest neighbor if that neighbor is doing better than $i$ at the end of the sequence (in monetary terms).
  \end{itemize}
\end{itemize}

@@

Our primary focus is on simulating the market in future scenarios. This is more relevant for policy design and more interesting theoretically due to the increasing divergence of global temperature values under the two future scenarios. The sensitivity analysis is based on a Latin hypercube sampling of 500 parameter sets from the following distributions \shortcite{beachkofski_improved_2002,carnell_lhs_2012}.

@@

We use the model to perform 10 full simulations for each of the 500 input parameter sets and average the 10 convergence scores. We conducted multiple simulations for the same parameter set because there is stochasticity. We then conducted a partial rank correlation coefficient analysis on the relationship between the input matrix, $X$, and the resulting simulated outcome vector of mean belief convergence scores, $y$ \shortcite{Marino2008,pujol_sensitivity:_2014,saltelli_sensitivity_2009}. Partial correlation computes the linear relationship between the part of the variation of $X_i$ and $y$ that are linearly independent of other $X_j$  ($j \neq i$).  The only difference between the partial correlation and the partial rank correlation that we use here is that the outcome variable $y$ is first ranked-transformed in order to allow us capture potentially non-linear relationships: every value, $y_s$, in the data set is replaced by a number 
$f(y_s) \equiv |\{y_r~|~ y_s > y_r\}|$, 
where $y_r$ are the observed values of $y$ in the data set.
